A Continuation-Based Solution\break of the Linearity Challenge

### 2025-12-29 paper/related.tex
final revision?

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for
languages with linear resources. One of the presented case studies is the
formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. 
%
Unlike the other formalisations we are discussing,
\citet{RouvoetPoulsenKrebbersVisser20} do not define a small-step semantics for
\GV which explains (in part) the small size of their development. However, since
their formalisation of \GV is intrinsically typed, the interpreter preserves
typing and therefore proves a form of subject reduction property.
%
The separating conjunction introduced in \Cref{sec:context-agda} and the typing
of the constructors for the representation of processes in
\Cref{sec:process-agda} have been adapted from this work of
\citet{RouvoetPoulsenKrebbersVisser20}.

@@

\citet{ZackonEtAl25} describe a typing context management technique where
channels are associated not just with a type but also with an element -- called
\emph{tag} -- of a given resource algebra that summarises the number of allowed
usages of a particular channel, including the possibility that the channel is
not available. This approach streamlines context splitting since contexts can be
treated in an essentially structural way, except for tags which are
conveninently combined using operations from the resource algebra.

@@

\Cref{tab:comparison} shows an overview of the aforementioned formalisations
(sorted by publication date) including our own. The first five columns identify
the calculus being formalised. We provide its reference paper, the prover in
which it is formalised and an acronym that gives an idea of the flavour of the
calculus. We also specify whether the calculus features \emph{cuts} (that is,
the combination of restriction and parallel composition corresponding with the
cut of linear logic) and continuations.
%
\CP~\cite{Wadler14,GayVasconcelos25} and \GV~\cite{GayVasconcelos10,Wadler14}
are well-known acronyms in the literature on session types. \SessionPI refers to
(variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants of) the
linear $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the
structural version of \CP introduced by \citet{SanoKavanaghPientka23} and
\Calculus is our calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may sometimes refer to different versions of the same calculus.
In particular, \GV in a logical setting is described by \citet{Wadler14} but its
first (non-logical) version is due to \citet{GayVasconcelos10}.

Concerning the use of continuations, the approaches based directly on the linear
$\pi$-calculus (into which sessions can be encoded) are marked with {\YES} and
those based on a calculus with native sessions are marked with {\NO}. The
calculus \SCP is marked with {\MAYBE} because, while not directly inspired to
the linear $\pi$-calculus, it makes use of explicit continuations for defining
the predicates that check the linear usage of channels. Finally, all the
approaches based on \GV are also marked with \MAYBE. In principle, \emph{none}
of these calculi makes use of continuations. However, \GV is designed in such a
way that each operation acting on a channel $s$ is a function that returns the
result of the operation (if present) \emph{along with the same channel $s$}. In
this way, the type of $s$ can be conveniently ``updated'' to take into account
the effect of the operation. As observed by \citet{Padovani17}, this semantics
of the communication primitives is virtually indistinguishable from one making
use of explicit continuation channels.

@@

The rightmost columns report the size (in kilobytes) of the formalisations as
rough (and possibly questionable) estimates of their complexity. Papers
describing formalisations typically use ``lines of code'' as a measure of
development effort, but the number of lines may be affected by code indentation
styles and syntactical constraints of the proof assistant being used. For this
reason, we have preferred to count the total number of characters after comments
have been removed and spaces have been squeezed.\footnote{Sequences of two or
more consecutive space-like characters are collapsed into a single space. The
squeezing is obtained by running the command \texttt{tr -s [:space:] file} on
Unix-like systems.}
%
The reported sizes account for the source code of the formalisations excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24}.
%
Some
formalisations~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,ZackonEtAl25}
define \emph{libraries} which can be reused in different contexts. In these
cases, the size of the library is reported separately next to the size of the
part of the development that uses it.

@@

Finally, it appears that the use of (explicit) continuations is related to the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, no matter if they
are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
ones are 263kb on average. This observation agrees with the considerations made
earlier in this paper concerning the difference between \emph{updating} bindings
versus \emph{removing and adding} bindings in a typing context. The use of
continuations enables a cleaner management of typing contexts since linear
channels are true ``use-once'' resources and there is no need to update their
type.

>>>>

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for
languages with linear resources. One of the presented case studies is the
formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. 
%
Unlike the other formalisations we are discussing,
\citet{RouvoetPoulsenKrebbersVisser20} do not define a small-step semantics for
\GV but their formalisation is intrinsically typed, hence the interpreter
preserves proves a form of subject reduction property.
%
The separating conjunction defined in \Cref{sec:context-agda} and the typing of
the constructors for the representation of processes in \Cref{sec:process-agda}
have been adapted from this work of \citet{RouvoetPoulsenKrebbersVisser20}.

@@

\citet{ZackonEtAl25} describe a typing context management technique where
channels are associated not just with a type but also with a \emph{tag}, that is
an element of a given resource algebra that summarises the number of allowed
usages of a particular channel, including the possibility that the channel is
not available. This approach streamlines context splitting since contexts can be
treated in an essentially structural way, except for tags which are
conveninently combined using operations from the resource algebra.

@@

\Cref{tab:comparison} shows an overview of the aforementioned formalisations
(sorted by publication date) including our own. The first five columns identify
the calculus being formalised. We provide its reference paper, the prover in
which it is formalised and an acronym that gives an idea of the flavour of the
calculus. We also specify whether the calculus features \emph{cuts} (that is,
the combination of restriction and parallel composition corresponding to the cut
of linear logic) and continuations.
%
\CP~\cite{Wadler14,GayVasconcelos25} and \GV~\cite{GayVasconcelos10,Wadler14}
are well-known acronyms in the literature on session types. \SessionPI refers to
(variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants of) the
linear $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the
structural version of \CP introduced by \citet{SanoKavanaghPientka23} and
\Calculus is our calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may sometimes refer to different versions of the same calculus.
In particular, \GV in a logical setting is described by \citet{Wadler14} but its
first (non-logical) version is due to \citet{GayVasconcelos10}.

Concerning the use of continuations, the approaches based directly on the linear
$\pi$-calculus (into which sessions can be encoded) are marked with {\YES} and
those based on a calculus with native sessions are marked with {\NO}. The
calculus \SCP is marked with {\MAYBE} because, while not directly inspired to
the linear $\pi$-calculus, it makes use of explicit continuations for defining
the predicates that check the linear usage of channels. Finally, all the
approaches based on \GV are also marked with \MAYBE. Officially, \emph{none} of
these calculi makes use of continuations. However, \GV is designed in such a way
that each operation acting on a channel $s$ is a function that returns the
result of the operation (if present) \emph{along with the same channel $s$}. In
this way, the type of $s$ can be conveniently ``updated'' to take into account
the effect of the operation. As observed by \citet{Padovani17}, this semantics
of the communication primitives is virtually indistinguishable from one making
use of explicit continuation channels.

@@

The rightmost columns report the size (in kilobytes) of the formalisations as
rough (and possibly questionable) estimates of their complexity. Papers
describing formalisations typically report the ``lines of code'' as a measure of
development effort, but the number of lines may be affected by code indentation
styles and syntactical constraints of the proof assistant being used. For this
reason, we have preferred to count the total number of characters after comments
have been removed and spaces have been squeezed.\footnote{Sequences of two or
more consecutive space-like characters are collapsed into a single space. The
squeezing is obtained by running the command \texttt{tr -s [:space:] file} on
Unix-like systems.}
%
The reported sizes account for the source code of the formalisations excluding
examples and any safety result, if present. We have excluded safety results
because their meaning varies widely across the formalisations and, except for
our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24}.
%
Some
formalisations~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,ZackonEtAl25}
define \emph{libraries} which can be reused in different contexts. In these
cases, the size of the library is reported separately next to the size of the
part of the development that uses it.

@@

Finally, it appears that the use of (explicit) continuations is related to the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, no matter if they
are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
ones are 263kb on average. At the very least, the use of continuations enables a
cleaner management of typing contexts since linear channels are true
``use-once'' resources and there is no need to update their type.
